\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-15 doc/paper-jss/sdmTMB-paper.Rnw
Editing

The process component of an \pkg{sdmTMB} model can be formed by any combination
of main (``fixed'') effects \(\bm{\beta}\) and offset \(O_{\bm{s},t}\), IID
(independent and identically distributed) random intercepts \(\alpha_g\),
spatial intercept random fields \(\omega_{\bm{s}}\), spatiotemporal intercept
random fields \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\),
and spatially varying effects \(\zeta_{\bm{s}}\) (Figure \ref{fig:diagram}).
These components are discussed in more detail via our examples below.

@@

By default, if spatiotemporal fields are included, they are assumed IID;
however, additional options allow them to be modelled as a random walk or
first-order autoregressive, AR(1), process (Figure \ref{fig:diagram}). Turning
off both spatial and spatiotemporal effects allows for comparison with standard
non-spatial GLMs or GLMMs. We include additional flexibility in specifying the
linear main effect matrix: covariates can be modelled as penalized smooth
functions (generalized additive models, GAMs) using the same \texttt{s()} syntax
as in \pkg{mgcv} \citep{wood2017a} (thereby allowing automatic selection of
smoother `wiggliness'), or can be modelled with threshold shapes: hockey-stick
models, \texttt{breakpt()}, \citep{barrowman2000} or logistic functions,
\texttt{logistic()}. \pkg{sdmTMB} can use many \texttt{mgcv::s()} and
\texttt{mgcv::t2()} smoothers including bivarate smoothers \texttt{s(x, y)},
smoothers varying by continuous or categorical variables \texttt{s(x1, by =
x2)}, cyclical smoothers \texttt{s(x, bs = "cc")}, and smoothers with
specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.

@@

<<pcod-head, echo=TRUE>>=
select(pcod, lat, lon, X, Y, depth, present) |>
  head()
@

@@

@@

<<dog-head-dat, echo=TRUE>>=
dat <- select(dogfish, lon = longitude, lat = latitude, year, 
  catch_weight, area_swept, depth) |> as_tibble()
head(dat)
@

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta or hurdle distribution: delta-gamma and
delta-lognormal \citep{aitchison1955}, and delta-Poisson-link-gamma and
delta-Poisson-link-lognormal \citep{thorson2018poisson}. The Poisson-link models
depart from the conventional delta models in treating the probability of
occurrence at a given location ($p_i$) as a Poisson process rather than
a Bernoulli one, $p_i = 1 - exp(a_i * n_i)$, where $n_i$ is the total number of
individuals available to be encountered, and $a_i$ is a measure of sampling
effort \citep{thorson2018poisson}. Note that we choose to omit the
spatiotemporal random fields from the binomial model component
(\texttt{spatiotemporal = list("off", "iid")}) because initial fits indicated
this random field was collapsing to zero indicating minimal changes in encounter
spatial distribution from year to year.

@@

We find that the Poisson-link delta-lognormal model is favoured with AIC. We
next test two additional model formulations: making the spatial correlation
isotropic (the default) instead of anisotropic, and structuring the positive
model component spatiotemporal random fields as AR(1) to allow spatiotemporal
hotspots to persist from year to year.

The AR(1) fields can be represented as TODO.

@@

We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy
is important here given the elongated shape of the continental shelf with
a rapid transition to deeper water. On inspecting the model output with
\texttt{summary()}, we notice that the spatial random field SD in the positive
model component has become very small once AR(1) spatiotemporal fields were
added. We therefore consider a model where these are turned off in the second
component of the delta model:

@@

We save our chosen model to the object \texttt{fit} to simplify subsequent code,
run the \texttt{sanity()} check (suppressed for space), and inspect
\texttt{summary()}:

@@

The output is more complex compared to our binomial spatial model. We now have
two model components, which are shown one after the other. Starting with the
binomial component, we have output from the smoother, which includes a linear
component (\texttt{slog\_depth}) and the standard deviation on the smoother
weights (\texttt{sds(log\_depth)}). The smoother summary follows the format used
in the \pkg{brms} package \citep{burkner2017}. Next, we have the time-varying
intercepts and information on our anisotropic spatial correlation. We then have
the second model component (lognormal) with a similar summary structure but with
the addition of a dispersion parameter for the lognormal, the AR(1) correlation
of the spatiotemporal random fields, and a spatiotemporal random field marginal
standard deviation. We can use \texttt{plot\_anisotropy()} to visually inspect
the anisotropy (Figure \ref{fig:dog-aniso}).

<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}. Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled. The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from zero.", out.width="4in">>=
plot_anisotropy(fit)
@

@@

Our prediction data frame is similar to the binomial spatial model, but includes
columns for the two delta model components (labelled with suffixes \texttt{1}
and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal
random effects. We can easily generate plots from this data frame (e.g., with
\texttt{ggplot2::geom\_raster()}). We suppress that code in this paper for
brevity.

@@

<<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density. Throughout, two example years are shown. \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects. The spatial random field is constant through time and represents static biotic or abiotic features not included as covariates (e.g. habitat). The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process. They represent temporal variability in the spatial patterning of Pacific Dogfish (resulting from movement, local changes in population densities, etc).", fig.width=9, out.width="6in">>=
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
@

We can visualize the conditional effect of the bottom depth smoother by
predicting across a sequence of depths and holding other variables at reference
values (Figure \ref{fig:dog-depth-plot}). Here, we pick the last year, specify
to include both delta model components (\texttt{model = NA}), omit the random
fields (\texttt{re\_form = NA}), and return standard errors (\texttt{se\_fit\
= TRUE}).

@@

<<dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
ggplot(pred_depth, aes(
  exp(log_depth), exp(est), 
  ymin = exp(est - 2 * est_se), 
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(x = "Depth (m)", y = "Density")
@

We can generate an area-weighted population index (e.g., a relative or absolute
index of abundance or biomass) that is independent of sampling locations by
predicting from the model on a grid covering the area of interest and summing
the predicted biomass with the \texttt{get\_index()} function (Figure
\ref{fig:dog-index-plot}). We supply the grid cell area (4
km\textsuperscript{2}) to the \texttt{area} argument and specify
\texttt{bias\_correct = TRUE} to enable a generic epsilon-bias correction
needed due to the non-linear transformation of the random effects (REF).

@@

<<dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(y = "Biomass", x = "Year")
@

@@

Snowy Owls (\emph{Bubo scandiacus}) breed on the arctic tundra and are irruptive
migrants, meaning that they appear across the mid-latitudes of North America in
much greater numbers in some winters than others. The reasons for this
interannual variation in the number of individuals migrating south are not well
understood but seem to be related to high abundances of food during the breeding
season and therefore sharp increases in breeding ground population densities
\citep{robillard2016}. The North Atlantic Oscillation Index (NAO) has been
linked to productivity of both owls and their prey in Europe \citep{millon2014}.
Because both productivity and the choice of wintering location could be
influenced by climate, we modelled a spatially varying effect of annual mean NAO
index on winter abundance across the southern boundary of their winter
distribution. We fit counts observed in North America during annual Christmas
Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011}
distribution, random intercepts for year, spatial and spatiotemporal random
fields, and a spatially varying coefficient associated with the NAO.

@@

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\alpha_g +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right),
\end{aligned}
\]

where three spatial fields are now estimated: the spatial intercept ($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and spatially varying NAO coefficients ($\zeta_{\bm{s}}$).

<<owl-fit, echo=TRUE, results = "hide">>=
mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owl <- sdmTMB(
  count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  mesh = mesh_snow,
  reml = TRUE,
  silent = FALSE
)
@

@@

@@

We can make a basic plot using the following code. A more elaborate
version is shown in Figure \ref{fig:owl-plot-fancy}.

@@

Overall, we find a weak average positive effect of annual mean NAO on overall
counts, but a southeast to northwest gradient in the intensity of the effect
(Figure \ref{fig:owl-plot-fancy}). This result is consistent with owls closest
to the Atlantic coast and those migrating the furthest south being the most
affected by NAO.

@@

We ran a simple speed comparison between \pkg{sdmTMB},
\proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE
spatial random field model to 1000 data points with Gaussian error across
a range of mesh resolutions (Figure \ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however,
\proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh
resolution than \pkg{sdmTMB}. \pkg{mgcv} was most affected by mesh resolution.
Our test was restricted to one core and default \proglang{R} algebra libraries;
all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (\pkg{openBLAS}:
\citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in
a \textasciitilde10\% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and
a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.

>>>>

The process component of an \pkg{sdmTMB} model can be formed by any combination
of main (``fixed'') effects \(\bm{\beta}\) and offset \(O_{\bm{s},t}\), IID
(independent and identically distributed) random intercepts \(\alpha_g\),
spatial intercept random fields \(\omega_{\bm{s}}\), spatiotemporal intercept
random fields \(\epsilon_{\bm{s},t}\), time-varying effects \(\bm{\gamma_t}\),
and spatially varying effects \(\zeta_{\bm{s}}\) (Figure~\ref{fig:diagram}).
These components are discussed in more detail via our examples below.

@@

By default, if spatiotemporal fields are included, they are assumed IID;
however, additional options allow them to be modelled as a random walk or
first-order autoregressive, AR(1), process (Figure~\ref{fig:diagram}). Turning
off both spatial and spatiotemporal effects allows for comparison with standard
non-spatial GLMs or GLMMs. We include additional flexibility in specifying the
linear main effect matrix: covariates can be modelled as penalized smooth
functions (generalized additive models, GAMs) using the same \texttt{s()} syntax
as in \pkg{mgcv} \citep{wood2017a} (thereby allowing automatic selection of
smoother `wiggliness'), or can be modelled with threshold shapes: hockey-stick
models, \texttt{breakpt()}, \citep{barrowman2000} or logistic functions,
\texttt{logistic()}. \pkg{sdmTMB} can use many \texttt{mgcv::s()} and
\texttt{mgcv::t2()} smoothers including bivarate smoothers \texttt{s(x, y)},
smoothers varying by continuous or categorical variables \texttt{s(x1, by =
x2)}, cyclical smoothers \texttt{s(x, bs = "cc")}, and smoothers with
specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.

@@

<<pcod-head, echo=TRUE>>=
select(pcod, lat, lon, X, Y, depth, present)
@

@@

\subsection{Adding UTM columns and creating a mesh}

@@

<<dog-head-dat, echo=TRUE>>=
dat <- select(dogfish, lon = longitude, lat = latitude, year, 
  catch_weight, area_swept, depth)
dat
@

@@

\subsection{Exploring delta-model alternative families}

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta or hurdle distribution: delta-gamma and
delta-lognormal \citep{aitchison1955}, and delta-Poisson-link-gamma and
delta-Poisson-link-lognormal \citep{thorson2018poisson}. The Poisson-link models
depart from conventional delta models in treating the probability of occurrence
at a given location ($p_i$) as a Poisson process rather than a Bernoulli one,
$p_i = 1 - \exp(a_i \cdot n_i)$, where $n_i$ is the total number of individuals
available to be encountered, and $a_i$ is a measure of sampling effort
\citep{thorson2018poisson}. Note that we choose to omit the spatiotemporal
random fields from the binomial model component (\texttt{spatiotemporal
= list("off", "iid")}) because initial fits indicated this random field was
collapsing to zero indicating minimal changes in encounter spatial distribution
from year to year.

@@

We find that the Poisson-link delta-lognormal model is favoured with AIC.

\subsection{Adding AR(1) random fields and comparing isotropic correlation}

We next test two additional model formulations: making the spatial correlation
isotropic (the default) instead of anisotropic, and structuring the positive
model component spatiotemporal random fields as AR(1) to allow spatiotemporal
hotspots to persist from year to year.

The AR(1) fields can be represented as:

@@

We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy
is important here given the elongated shape of the continental shelf with
a rapid transition to deeper water. We can use \texttt{plot\_anisotropy()} to
visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).

<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}. Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled. The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).", out.width="4in">>=
plot_anisotropy(fit)
@

On inspecting the model output with
\texttt{summary()}, we notice that the spatial random field SD in the positive
model component has become very small once AR(1) spatiotemporal fields were
added. We therefore consider a model where these are turned off in the second
component of the delta model and choose this model.

@@

We save our chosen model to the object \texttt{fit} to simplify subsequent code,
run the \texttt{sanity()} check (suppressed for brevity), and inspect
\texttt{summary()}:

@@

The output is more complex compared to our binomial spatial model. We now have
two model components, which are shown one after the other. Starting with the
binomial component, we have output from the smoother, which includes a linear
component (\texttt{slog\_depth}) and the standard deviation on the smoother
weights (\texttt{sds(log\_depth)}). The smoother summary follows the format used
in the \pkg{brms} package \citep{burkner2017}. Next, we have the time-varying
intercepts and information on our anisotropic spatial correlation. We then have
the second model component (lognormal) with a similar summary structure but with
the addition of a dispersion parameter for the lognormal, the AR(1) correlation
of the spatiotemporal random fields, and a spatiotemporal random field marginal
standard deviation.

\subsection{Visualizing model predictions}

@@

Our prediction data frame is similar to the binomial spatial model, but includes
columns for the two delta model components (labelled with suffixes \texttt{1}
and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal
random effects. We can easily generate plots from this data frame using code
similar to in our spatial example with Pacific Cod. We suppress that code for
brevity.

@@

<<dog-wcvi-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density. Throughout, two example years are shown. \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects. The spatial random field is constant through time and represents static biotic or abiotic features not included as covariates (e.g. habitat). The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process. They represent temporal variability in the spatial patterning of Pacific Spiny Dogfish (e.g., resulting from movement or local changes in population density).", fig.width=9, out.width="6in">>=
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
@

We can visualize the conditional effect of the bottom depth smoother by
predicting across a sequence of depths and holding other variables at reference
values (Figure~\ref{fig:dog-depth-plot}). Here, we pick the last year, specify
to include both delta model components (\texttt{model = NA}), omit the random
fields (\texttt{re\_form = NA}), and return standard errors (\texttt{se\_fit\
= TRUE}).

@@

<<dog-depth-plot, echo=FALSE, fig.cap="The conditional effect of ocean bottom depth on Pacific Spiny Dogfish density. The line and shaded ribbon represent the mean and 95\\% confidence interval, respectively. Other fixed effects are held at constant values and the random fields are omitted from the prediction.", out.width="4in">>=
ggplot(pred_depth, aes(
  exp(log_depth), exp(est), 
  ymin = exp(est - 2 * est_se), 
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(x = "Depth (m)", y = "Density")
@

\subsection{Calculating an area-weighted index}

We can generate an area-weighted population index (e.g., a relative or absolute
index of abundance or biomass) that is independent of sampling locations by
predicting from the model on a grid covering the area of interest and summing
the predicted biomass with the \texttt{get\_index()} function (Figure
\ref{fig:dog-index-plot}). We supply the grid cell area (4
km\textsuperscript{2}) to the \texttt{area} argument and specify
\texttt{bias\_correct = TRUE} to enable a bias correction needed due to the
non-linear transformation of the random effects \citep{thorson2016bias}.

@@

<<dog-index-plot, echo=FALSE, fig.cap="Area-weighted index of relative biomass over time for Pacific Spiny Dogfish. Dots and line segments represent means and 95\\% confidence intervals.", out.width="4in">>=
ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(y = "Biomass", x = "Year")
@

@@

Snowy Owls (\emph{Bubo scandiacus}) breed on the arctic tundra and are irruptive
migrants, meaning that they appear across the mid-latitudes of North America in
much greater numbers in some winters than others. The reasons for this
interannual variation in the number of individuals migrating south are not well
understood but seem to be related to high abundances of food during the breeding
season and therefore sharp increases in breeding ground population densities
\citep{robillard2016}. The North Atlantic Oscillation Index (NAO) has been
linked to productivity of both owls and their prey in Europe \citep{millon2014}.
Because both productivity and the choice of wintering location could be
influenced by climate, we modelled a spatially varying effect of annual mean NAO
index on winter abundance across the southern boundary of their winter
distribution.

Our data are read in from supplementary data and contain columns for UTM spatial coordinates, year, year as a factor, and the annual NAO value:

<<owl-data-head, echo=TRUE>>=
select(snow, X, Y, year, year_f, nao)
@

We fit counts observed in North America during annual Christmas
Bird Counts \citep{cbc} using a negative binomial \citep[NB2,][]{hilbe2011}
distribution, random intercepts for year, spatial and spatiotemporal random
fields, and a spatially varying coefficient associated with the NAO.

@@

\[
\begin{aligned}
y_{\bm{s},t} &\sim \operatorname{NB2} 
  \left(
  \mu_{\bm{s},t}, 
  \phi 
  \right),\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\alpha_t +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right),
\end{aligned}
\]

where three types of random fields are now estimated: the spatial intercept
($\omega_{\bm{s}}$), spatiotemporal variation ($\epsilon_{\bm{s},t}$), and
spatially varying NAO coefficients ($\zeta_{\bm{s}}$). The NB2 distribution is
specified with a mean $\mu_{\bm{s},t}$ and size parameter $\phi$. The
observation variance scales quadratically with the mean: $\mathrm{Var}[y] = \mu
+ \mu^2 / \phi$ \citep{hilbe2011}. The $\alpha_t$ represent IID random
intercepts by year: $\alpha_t \sim \operatorname{Normal} \left(0,
\sigma_\alpha^2 \right)$.

<<owl-fit, echo=TRUE, results = "hide">>=
mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owl <- sdmTMB(
  count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  mesh = mesh_snow,
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  reml = TRUE,
  silent = FALSE
)
@

@@

In addition to the output seen for other models, we now have a section for
random intercepts and a standard deviation for the spatially varying coefficient
random field. Given our model specification, all random fields are sharing
a single Matérn range.

\subsection{Extracting the spatially varying coefficient effects}

@@

We can make a basic plot using the following code. A more elaborate
version is shown in Figure~\ref{fig:owl-plot-fancy}.

@@

Overall, we find a weak average positive effect of annual mean NAO on overall
counts, but a southeast to northwest gradient in the intensity of the effect
(Figure~\ref{fig:owl-plot-fancy}). This result is consistent with owls closest
to the Atlantic coast and those migrating the furthest south being the most
affected by NAO.

@@

We ran a simple speed comparison between \pkg{sdmTMB},
\proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE
spatial random field model to 1000 data points with Gaussian error across
a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however,
\proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh
resolution than \pkg{sdmTMB}. \pkg{mgcv} was most affected by mesh resolution.
Our test was restricted to one core and default \proglang{R} algebra libraries;
all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (\pkg{openBLAS}:
\citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in
a \textasciitilde10\% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and
a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.
